---
title: "Time Series Assignment 4"
output: html_document
date: "2025-02-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(fpp)
library(fpp2)
library(TTR)

data <- read.csv("Unemployment_Canada_1976_present.csv")
data$REF_DATE <- as.Date(paste0(data$REF_DATE, "-01"))  # Convert to Date

##weighted average
datafiltered <- data %>%
  group_by(REF_DATE) %>%
  summarize(Unemployment_Rate = weighted.mean(Unemployment.rate, Labour.force, na.rm=TRUE))

tsdata <- ts(datafiltered$Unemployment_Rate, start=c(1976,1), frequency=12)

```

```{r}
plot(tsdata, main="Unemployment Rate in Canada", ylab="Rate", xlab="Year")
Acf(tsdata, main="ACF of Unemployment Rate")

etsforecast <- ets(tsdata)
plot(ets_forecast)
title("ETS Decomposition of Unemployment Rate")
attributes(ets_forecast)
etsforecast$mse
```
## According to the ACF plot, there are no consistent spikes at the intervals of 12, showing that there is no significant seasonality present. However, is a downward trend showing in the spikes. The decomposition also shows that there is a trend but no seasonality. 

```{r}
holtforecast <- holt(tsdata, h=5, beta=0.2)
plot(holtforecast, main="Holt's Exponential Smoothing")

meanforecast <- meanf(tsdata, h=5)
naiveforecast <- naive(tsdata, h=5)
rwf_forecast <- rwf(tsdata, h=5)
rwf_forecast_drift <- rwf(tsdata, h=5, drift=TRUE)
seasonalnaiveforecast <- snaive(tsdata, h=5)

plot(tsdata, main="Forecast Comparisons", ylab="Unemployment Rate", xlab="Year", col="black")
lines(meanforecast$mean, col="blue", lwd=2)
lines(naiveforecast$mean, col="red", lwd=3)  # Increase line width
lines(rwf_forecast$mean, col="green", lwd=2)
lines(seasonalnaiveforecast$mean, col="purple", lwd=2)
lines(holtforecast$mean, col="orange", lwd=2)
legend("topleft", legend=c("Mean", "Naive", "Random Walk", "Seasonal Naive", "Holt"),
       col=c("blue", "red", "green", "purple", "orange"), lty=1, lwd=3)



```

```{r}
tsaccuracy <- data.frame(
  Mean = accuracy(meanforecast)[2],
  Naive = accuracy(naiveforecast)[2],
  Seasonal_Naive = accuracy(seasonalnaiveforecast)[2],
  Holt = accuracy(holtforecast)[2]
)


tsaccuracy
```
## As we can see by the errors, the Naive method is the best which could be explained by recent stability in uemployment, closely followed by Holt, which I used over Holt-Winters due to the absence of seasonality. In the long-term, the Holt-model could be a more suitable model to use. 
